Research Spike: Evaluation of Using a Single Mnemonic both for Cryptocurrency and Identity Wallets

### 2022-01-10 prism-backend/research/single-mnemonic/bip32-etal.tex
Changes based on Tony's comments.

Hierarchical Deterministic Wallets (HD wallets, for short) in the cryptocurrency
domain derive from the BIP32 specification%
\footnote{\url{https://github.com/bitcoin/bips/blob/master/bip-0032.mediawiki}.
  Last access, December 13th, 2021.}
The motivation behind creating this type of wallets is clear if one thinks about
how typical cryptocurrency systems work: each address is associated with a key
pair and, in order to achieve high security and privacy levels, it is best to
limit the reusage of addresses (thus, key pairs). Therefore, a too high number
of needed key pairs is quickly reached. If each key is generated independently
at random, management becomes too complex -- especially, if several devices
are expected to be synchronised. As a solution to this challenge, a hierachical
tree-based data structure was proposed, in which each parent node can be used to
derive multiple child nodes, deterministically. This approach seems also useful
to foster some sort of separation -- i.e., create sub-trees that are
(computationally) independent from one another, in the sense that a compromise
in one does not lead to a compromise in the other. Typically, the root node
of the tree is derived from a seed encoded as a mnemonic. BIP39%
\footnote{\url{https://github.com/bitcoin/bips/blob/master/bip-0039.mediawiki}.
  Last access, December 13th, 2021.} is the main specification for this purpose.

@@

\subsection{Main Specifications}

While the main specification is BIP32, it essentially gives the basic guidelines
to create HD wallets. To avoid ambiguity leading to incompatible
implementations, BIP43%
\footnote{\url{https://github.com/bitcoin/bips/blob/master/bip-0043.mediawiki}.
  Last access, December 13th, 2021.} and BIP44%
\footnote{\url{https://github.com/bitcoin/bips/blob/master/bip-0044.mediawiki}.
  Last access, December 13th, 2021.} specify how to organize HD wallets so
that the resulting tree structure enables the most common use cases (which, at
least initially, focus on the cryptocurrency scenario). In addition, the
specification that defines how to encode (pseudo-)random bitstrings as
mnemonic is given in BIP39.

\subsection{Main Concepts}

@@

@@

\paragraph{Layer 1: \emph{isolated} subtrees per \texttt{purpose}} %
The first layer of a BIP44-compatible HD wallet is composed of up to $2^{31}$
nodes of hardened childs: i.e., layer 1 is composed of childs $(1,2^{31})$
to $(1,2^{62}-1)$, where the second element in the pair is the $i$ value used
in the child derivation processes, and denotes for what ``purpose'' the
descendant keys will be used -- this roughly translates to the system (e.g.,
Bitcoin, Cardano... or even non-blockchain systems, in theory).

@@

For BIP44-compatible wallets, the expected index is $44$. Cardano, since
the Shelley era, uses $1852$%
\footnote{\url{https://cips.cardano.org/cips/cip1852/}. Last access, December
  14th, 2021.}. Note that, since these are hardened indexes, the encoding
rules translate this into $1852+2^{31}$ (resp. $44+2^{31}$ for BIP44); see
BIP44 for more details.

@@

Although not mandatory, each coin is expected to have an assigned coin type
index. The complete list of reserved indexes is available in SLIP44%
\footnote{\url{https://github.com/satoshilabs/slips/blob/master/slip-0044.md}.
  Last access, December 14th, 2021.}.

@@

\paragraph{Layer 4: \emph{non-isolated} subtrees per \texttt{change}}
According to BIP32, the fourth layer can be composed of up to  $2^{31}$ nodes of
non-hardened childs: i.e., layer 4 can be composed of childs $(4,0)$ to
$(4,2^{31}-1)$ per each child node of layer 3. However, in BIP44, only
non-hardened childs 0 and 1 are used: i.e., each node of layer 3 will only have
non-hardened childs $(4,0)$ and $(4,1)$. This layer is used to differentiate
whether the descendant nodes will be used as a change addresses, or not. I.e.,
child $(4,0)$ will \emph{not} be used for receiving change in payments, and
therefore is expected to be associated to an address that may be published
outside of the wallet; while child $(4,1)$ is expected to be used for receiving
change in payments.

Note that, since the 4th layer is not derived in hardened mode, if the private
key of the $(4,0)$ child of some layer 3 node is compromised along with the
extended public key of the node at layer 3, then the entire subtree of that
layer 3 node will be compromised (but not the subtrees of other layer 3 nodes,
as layer 3 is derived in hardened mode). That is, all the addresses (change or
not) of an specific account of some cryptocurrency, will be compromised.

>>>>

@@

\subsection{Main Specifications: A Brief History of BIPs for Wallets}

BIPs, or Bitcoin Improvement Proposals, are ``standards'' in the Bitcoin domain
that specify how concrete components of the ecosystem work and/or should be
implemented. Being Bitcoin the first blockchain, many others import these
standards.

In 2013, BIP32 was published. It defines an approach to HD wallets
using a tree structure in which each node can be associated to a key pair. It
also defines two ways to derive child nodes from parent nodes: one that allows,
given a public key of a node, to derive the public and private keys of all its
descendants; and another one that requires the private key for derivation of
its child nodes. Based on this child derivation processes, it also specifies a
basic strategy for building HD structures for payments.

However, although BIP32 does specify a basic structure, it does not mandate that
it has to be followed. According to BIP43%
\footnote{\url{https://github.com/bitcoin/bips/blob/master/bip-0043.mediawiki}.
  Last access, December 13th, 2021.}, published in 2014, this led to a set of
self-claimed BIP32-compliant wallets not to be compliant between themselves.
BIP43 requires that the first level of child derivation (directly from the
root) be dedicated to define a ``purpose''. BIP44%
\footnote{\url{https://github.com/bitcoin/bips/blob/master/bip-0044.mediawiki}.
  Last access, December 13th, 2021.}, published simultaneously to BIP43, further
refines this, by establishing 6 different levels (which we introduce in
subsequent sections). Thus, BIP44 is an specialization of BIP43, which is itself
an specialization of BIP32.

Somehow orthoghonally to these, BIP39 defines how to encode random binary seeds
into mnemonic sentences, and back. In this manner, the root node of a BIP44
wallet is decoded into a bitstring, from a mnemonic, using BIP39. Then, this
root node and its child nodes are computed as defined in BIP44. \tabref{tab:bips}
summarizes the mentioned standards.

\begin{table}[ht!]
  \centering
  \scalebox{0.95}{
    \begin{tabular}{|c|c|l|}
      \hline
      \bf BIP & \bf Year & \bf Description \\
      \hline
      BIP32 & 2013 & Basic structure and derivation rules for HD wallets. \\
      BIP43 & 2014 & Definition of ``purpose'' level. \\
      BIP44 & 2014 & Definition of 6-level structure (root, purpose, coin type,
                     account, change, address.) \\
      BIP32 & 2013 & Encoding/decoding of bitstrings to/from mnemonic sentences. \\
      \hline
    \end{tabular}
  }
  \caption{Summary of BIPs for HD wallets.}
  \label{tab:bips}
\end{table}

HD wallets that follow the BIP44 specification have to define a value of 44
(with hardened derivation; see next) at the ``purpose'' layer. Wallets following
a different structure, but still compatible with BIP32, can (should?) define
their own code, for the sake of ensuring compatibility across implementations.
For instance, Cardano defines the code 1852 (again, hardened derivation), as
it has some minor variations with respect to BIP44. There are several well-known
wallets that follow this specification. For instance, Trezor%
\footnote{\url{https://docs.trezor.io/trezor-firmware/misc/coins-bip44-paths.html}.
  Last access, January 10th, 2022.}, or Ledger%
\footnote{\url{https://support.ledger.com/hc/en-us/articles/4404388633489-Export-your-accounts?docs=true}.
Last access, January 10th, 2022.}.

Additionally, Satoshi Labs SLIP44 maintains a registry of the different
cryptocurrencies that have ``registered'' codes for wallet specifications
that follow BIP44. That is, this registry contains codes for the ``coin type''
level of BIP44. Examples of registered coins include Bitcoin mainnet and testnet,
Litecoin, Cardano, Ether, and many others (currently, there are several
hundreds).

% While the main specification is BIP32, it essentially gives the basic guidelines
% to create HD wallets. To avoid ambiguity leading to incompatible
% implementations, BIP43 and BIP44 specify how to organize HD wallets so
% that the resulting tree structure enables the most common use cases (which, at
% least initially, focus on the cryptocurrency scenario). In addition, the
% specification that defines how to encode (pseudo-)random bitstrings as
% mnemonic is given in BIP39.

\subsection{Main Concepts}
\label{ssec:mainhdconcepts}

@@

\paragraph{Path notation.} As can be seen in \figref{fig:bip44-tree}, the
hierarchical structure of an HD wallet is defined in layers (or levels), where
different indexes are used to derive the cryptographic values associated to
different nodes in a same layer, with the exception of the root node (at layer
0), which is frequently just denoted with an \texttt{m}, after master seed.
Thus, for instance, paths  \texttt{m/1} and \texttt{m/2} denote a tree with two
layer 1 nodes computed with index \texttt{1} and index \texttt{2},
respectively. If a child node index is \emph{primed}, it means that hardened
derivation is used (whereas the absence of a prime means non-hardened
derivation). More specifically, since BIP44 wallets have up to layer 6, paths
look like \texttt{m/a'/b'/c'/d'/e/f}. Note that layers 1 to 3 are derived in
hardened mode, and layers 4 and 5 are derived in non-hardened mode.

@@

\paragraph{Layer 1: \emph{isolated} subtrees per \texttt{purpose}} %
The first layer of a BIP44-compatible HD wallet is composed of up to $2^{31}$
nodes of hardened childs: i.e., layer 1 is composed of childs $(1,2^{31})$
to $(1,2^{62}-1)$, where the second element in the pair is the index $i$,
whose value is used in the child derivation processes, and denotes for what
``purpose'' the descendant keys will be used -- this roughly translates to the
system (e.g., Bitcoin, Cardano... or even non-blockchain systems, in theory).

@@

For BIP44-compatible wallets, the expected value for the $i$ index is $44$, so
all paths begin with \texttt{m/44'}. Cardano, since the Shelley era, uses index
$1852$ (and thus, Cardano paths begin with \texttt{m/1852'}%
\footnote{\url{https://cips.cardano.org/cips/cip1852/}. Last access, December
  14th, 2021.}. Note that, since these are hardened indexes, the encoding
rules translate this into $1852+2^{31}$ (resp. $44+2^{31}$ for BIP44); see
BIP44 for more details.

@@

Although not mandatory, each coin is expected to have an assigned coin type
index. The complete list of reserved indexes is available in SLIP44%
\footnote{\url{https://github.com/satoshilabs/slips/blob/master/slip-0044.md}.
  Last access, December 14th, 2021.}. For instance, Cardano's ADA has an index
of 1815 in hardened mode -- thus, its path prefix is \texttt{m/1852'/1815'}.

@@

\paragraph{Layer 4: \emph{non-isolated} subtrees per \texttt{change}}
According to BIP32, the fourth layer can be composed of up to  $2^{31}$ nodes of
non-hardened childs: i.e., layer 4 can be composed of childs $(4,0)$ to
$(4,2^{31}-1)$ per each child node of layer 3. However, in BIP44, only
non-hardened childs 0 and 1 are used: i.e., each node of layer 3 will only have
non-hardened childs $(4,0)$ and $(4,1)$. This layer is used to differentiate
whether the descendant nodes will be used as change addresses, or not. I.e.,
child $(4,0)$ will \emph{not} be used for receiving change in payments, and
therefore is expected to be associated to an address that may be published
outside of the wallet; while child $(4,1)$ is expected to be used for receiving
change in payments, and will typically not be published outside of the
transaction.

Note that, since nodes in layer 4 are not derived in hardened mode, if the private
key of a layer 4 node is compromised, along with the extended public key of its
layer 3 parent, then the entire subtree of that layer 3 node will be compromised
(but not the subtrees of other layer 3 nodes, as layer 3 is derived in hardened
mode). For example, take the derivation path of a layer 4 node to be
\texttt{m/x'/y'/z'/0}. If the private key of \texttt{m/x'/y'/z'/0} is leaked to
an attacker, who also gains access to the \emph{extended public} key (i.e.,
the public key and chaincode) of \texttt{m/x'/y'/z'}, then the attacker will
be able to compute the extended private (and public) keys of the sibling node
\texttt{m/x'/y'/z'/1}. This is precisely the attack explained in the last
paragraph of \secref{ssec:mainhdconcepts}.

### 2022-01-10 prism-backend/research/single-mnemonic/introduction.tex
Changes based on Tony's comments.

>>>>

Hierarchical Deterministic Wallets (HD wallets, for short) in the cryptocurrency
domain derive from the BIP32 specification%
\footnote{\url{https://github.com/bitcoin/bips/blob/master/bip-0032.mediawiki}.
  Last access, December 13th, 2021.}
The motivation behind creating this type of wallets is clear if one thinks about
how typical cryptocurrency systems work: each address is associated with a key
pair and, in order to achieve high security and privacy levels, it is best to
limit the reusage of addresses (thus, key pairs). Therefore, a too high number
of needed key pairs is quickly reached. If each key is generated independently
at random, management becomes too complex -- especially, if several devices
are expected to be synchronised. As a solution to this challenge, a hierachical
tree-based data structure was proposed, in which each parent node can be used to
derive multiple child nodes, deterministically. This approach seems also useful
to foster some sort of separation -- i.e., create sub-trees that are
(computationally) independent from one another, in the sense that a compromise
in one does not lead to a compromise in the other. Typically, the root node
of the tree is derived from a seed encoded as a mnemonic. BIP39%
\footnote{\url{https://github.com/bitcoin/bips/blob/master/bip-0039.mediawiki}.
  Last access, December 13th, 2021.} is the main specification for this latter
purpose.

\subsection{Preliminaries}

In the sequel, the processes that compose an (BIP44-compliant) HD wallet are
described, including high-level cryptographic algorithms. For the sake of
readability of non-cryptography savvy audience, we give informal descriptions of
the cryptographic concepts and schemes mentioned next.

\paragraph{Signature schemes.} A signature scheme uses an asymmetric key pair,
composed of a public key and a private key (also frequently referred to as
verification key and signing key, respectively). Intuitively, the owner of the
signing key can produce a digital signature over arbitrary messages, which can
be verified by anyone with the verification key, which is frequently somehow
made publicly accessible. Secure signature schemes ensure that only the owner
of the signing key can produce signatures verifiable with the corresponding
verifying key. In the HD wallets studied next, typical signing algorithms are
ECDSA and EdDSA.

\paragraph{HMACs and HKDFs.} A Hash-based Message Authentication Code scheme is
a concrete approach to build MAC functions, using cryptographic hashes as main
building block. A MAC function is a kind of symmetric equivalent to digital
signature -- i.e., it relies on a shared secret rather than on an asymmetric key
pair. HMACs have been proved to be secure pseudo random number
generators \cite{bck96}, which is a useful building block to build secure HKDFs
\cite{kraw10}, or Hash-based Key Derivation Functions. Specifically, an HKDF, as
defined in \cite{kraw10}, can be applied to produce secure cryptographic keys
from low entropy sources, by applying a two-phase process of randomness
extraction and expansion (the extract-expand approach). In summary, given a
(possibly low-entropy) bitstring, an HKDF produces an arbitrarily long pseudo
random bitstring that is fit for cryptographic purposes.
